# Connect to an IDBac Database


Load the necessary packages for this tutorial:
```{r eval=TRUE, message=FALSE, warning=FALSE, include=TRUE}
library(IDBacApp)
```



## Connect to IDBac database

Here we will make a connection to the database. 

Under the hood IDBac uses the `{pool}` package and some IDBac functions expect a `pool` object and thus it is the preferred way to handle IDBac databases from within R.


```{r}
example_pool <- IDBacApp::idbac_connect(fileName = "idbac_experiment_file",
                                        filePath = here::here("data",
                                                              "example_data"))

```


Notice that this isn't a pool object, it's a list (this is for a reason- it helps the IDBac Shiny app make multiple connections).

```{r}
class(example_pool)
```

To get our pool object let's pull it out of the list.
```{r}
# example_pool$idbac_experiment_file would also work 
example_pool <- example_pool[[1]]
class(example_pool)
```

You should now be connected to an IDBac experiment/database!
